---
title: "Accessing and Analyzing RCSB PDB Data with rPDBapi"
author: "Selcuk Korkmaz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Accessing and Analyzing RCSB PDB Data with rPDBapi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)

if (!requireNamespace("rPDBapi", quietly = TRUE) &&
    requireNamespace("pkgload", quietly = TRUE)) {
  package_root <- if (file.exists("DESCRIPTION")) {
    "."
  } else if (file.exists(file.path("..", "DESCRIPTION"))) {
    ".."
  } else {
    NA_character_
  }
  
  if (!is.na(package_root)) {
    pkgload::load_all(package_root, export_all = FALSE, quiet = TRUE)
  }
}

library(rPDBapi)
library(dplyr)

live_examples <- identical(tolower(Sys.getenv("RPDBAPI_RUN_LIVE", "false")), "true")
have_r3dmol <- requireNamespace("r3dmol", quietly = TRUE)
```

# Introduction

The Protein Data Bank (PDB) is the primary global archive for experimentally
determined three-dimensional structures of biological macromolecules. The RCSB
PDB exposes these data through programmatic interfaces that support search,
metadata retrieval, coordinate download, and access to assembly- and
entity-level annotations. For structural bioinformatics, these APIs make it
possible to move from a biological question to a reproducible computational
workflow without manual browsing of the PDB website.

`rPDBapi` provides an R interface to these services. It combines search helpers,
operator-based query construction, metadata retrieval, GraphQL-based data
fetching, and structure download utilities in a form that fits naturally into
R-based data analysis pipelines.

This vignette is written for users who want to retrieve and analyze PDB data
directly in R. The examples focus on protein kinase structures because kinases
are biologically important, structurally diverse, and common targets in
drug-discovery workflows.

```{r live-note, results = "asis"}
if (!live_examples) {
  cat(
    "This vignette contains live API examples. ",
    "To execute them during rendering, set `RPDBAPI_RUN_LIVE=true` before knitting. ",
    "When the flag is not set, the code is still shown but network-dependent chunks are not evaluated.\n",
    sep = ""
  )
}
```

The note above makes the vignette safe to build in offline or CRAN-like
environments while keeping all examples directly executable in a live session.
This is especially important for API-driven packages, where remote services are
part of the normal workflow but should not make documentation brittle.

# Installation and Setup

```{r installation, eval = FALSE}
install.packages("rPDBapi")

# Development version
remotes::install_github("selcukorkmaz/rPDBapi")
```

The package can be installed from CRAN or from the development repository. The
development version is useful when you want the newest API mappings, tests, and
documentation updates. In a package vignette, installation code is normally
shown but not run, because users execute it once interactively rather than
during every render.

```{r libraries}
library(rPDBapi)
library(dplyr)
```

This chunk loads the package and `dplyr`, which we will use for simple
tabulation and ranking. Most structural bioinformatics workflows combine API
access with data manipulation, so it is useful to establish that pattern from
the beginning.

# Why Access the PDB from R?

Programmatic PDB access is valuable when you need to:
  
  - search large structural collections reproducibly
- retrieve metadata for many entries at once
- move from identifiers to tidy analysis tables
- combine structure data with statistics, visualization, and modeling tools in R

In practice, this means that a question such as "Which high-resolution protein
kinase structures are available, what organisms do they come from, and what do
their assemblies look like?" can be answered in one analysis script instead of
through manual web browsing.

# rPDBapi Capabilities

At a high level, the package supports seven related tasks:
  
  1. Search the archive with simple or structured queries.
2. Retrieve entry-, entity-, assembly-, or chemical-component metadata.
3. Discover and validate retrievable fields before issuing a request.
4. Normalize, infer, and build identifiers across PDB record levels.
5. Download coordinate files and parse them into R objects.
6. Convert nested API responses into analysis-ready data frames or richer typed objects.
7. Scale retrieval with batch fetching, cache management, provenance, and analysis helpers.

The current package version also hardens return contracts and error classes.
That matters when `rPDBapi` is embedded in larger pipelines, because downstream
code can now make stronger assumptions about object classes, identifier
formats, and failure modes.

# Package Feature Map

`rPDBapi` exports functions that fall into nine practical groups:
  
  - Search helpers: `query_search()`, `perform_search()`
- Search operators: `DefaultOperator()`, `ExactMatchOperator()`,
`InOperator()`, `ContainsWordsOperator()`, `ContainsPhraseOperator()`,
`ComparisonOperator()`, `RangeOperator()`, `ExistsOperator()`,
`SequenceOperator()`, `SeqMotifOperator()`, `StructureOperator()`,
`ChemicalOperator()`
- Query composition helpers: `QueryNode()`, `QueryGroup()`,
`RequestOptions()`, `infer_search_service()`, `ScoredResult()`
- Identifier helpers: `infer_id_type()`, `parse_rcsb_id()`,
`build_entry_id()`, `build_assembly_id()`, `build_entity_id()`,
`build_instance_id()`
- Metadata retrieval: `data_fetcher()`, `fetch_data()`, `generate_json_query()`,
`get_info()`, `find_results()`, `find_papers()`, `describe_chemical()`,
`get_fasta_from_rcsb_entry()`
- Schema-aware retrieval helpers: `list_rcsb_fields()`,
`search_rcsb_fields()`, `validate_properties()`, `add_property()`
- Batch, cache, and provenance helpers: `data_fetcher_batch()`,
`cache_info()`, `clear_rpdbapi_cache()`
- Structure and file retrieval: `get_pdb_file()`, `get_pdb_api_url()`
- Rich objects and analysis helpers: `as_rpdb_entry()`,
`as_rpdb_assembly()`, `as_rpdb_polymer_entity()`,
`as_rpdb_chemical_component()`, `as_rpdb_structure()`,
`summarize_entries()`, `summarize_assemblies()`,
`extract_taxonomy_table()`, `extract_ligand_table()`,
`extract_calpha_coordinates()`, `join_structure_sequence()`
- Low-level API and parsing utilities: `send_api_request()`,
`handle_api_errors()`, `parse_response()`, `search_graphql()`,
`return_data_as_dataframe()`

The main workflow only needs a subset of these functions, but the full package
is designed as a layered interface. High-level helpers are convenient for
routine work, while low-level helpers make it possible to debug requests, build
custom workflows, or extend the package into larger pipelines.

# Core Concepts in the RCSB PDB API

Before starting with code, it helps to distinguish a few PDB concepts:
  
  - `ENTRY`: a PDB deposition such as `4HHB`
- `ASSEMBLY`: a biological assembly within an entry, such as `4HHB-1`
- `POLYMER_ENTITY`: a unique macromolecular entity within an entry, such as a protein chain definition
- `ATTRIBUTE`: a searchable or retrievable field, such as `exptl.method` or `rcsb_entry_info.molecular_weight`

`rPDBapi` mirrors these levels. Search functions return identifiers at the
appropriate level, and metadata functions use those identifiers to fetch the
corresponding records.

```{r concepts}
kinase_full_text <- DefaultOperator("protein kinase")
high_resolution <- RangeOperator(
  attribute = "rcsb_entry_info.resolution_combined",
  from_value = 0,
  to_value = 2.5
)
xray_method <- ExactMatchOperator(
  attribute = "exptl.method",
  value = "X-RAY DIFFRACTION"
)

kinase_query <- QueryGroup(
  queries = list(kinase_full_text, xray_method, high_resolution),
  logical_operator = "AND"
)

kinase_query
```

This code builds a structured query object without contacting the API. The
query says: search for records related to "protein kinase", require
X-ray diffraction as the experimental method, and restrict the results to
structures with a reported resolution of 2.5 angstroms or better. This is a
useful pattern because it separates biological intent from the mechanics of the
HTTP request.

```{r request-options}
search_controls <- RequestOptions(
  result_start_index = 0,
  num_results = 10,
  sort_by = "score",
  desc = TRUE
)

search_controls
```

`RequestOptions()` defines how many hits to return and how to sort them. In
other words, the query object describes what you want, and the request options
describe how you want it delivered. That distinction matters when you are
iterating over result pages or creating reproducible subsets for downstream
analysis.

```{r identifier-helpers}
example_ids <- c("4HHB", "4HHB-1", "4HHB_1", "4HHB.A", "ATP")

dplyr::tibble(
  id = example_ids,
  inferred_type = infer_id_type(example_ids)
)

parse_rcsb_id("4HHB-1")
build_entry_id(" 4HHB ")
build_assembly_id("4HHB", 1)
build_entity_id("4HHB", 1)
build_instance_id("4HHB", "A")
```

These helpers make identifier handling explicit. They are useful when a
workflow moves across entry-, assembly-, entity-, and chain-level retrieval,
because the required identifier syntax changes with the biological level. In
practice, the helpers reduce ad hoc string handling and make it easier to write
validation checks before a request is sent.

# Workflow 1: Simple Search for Kinase Structures

```{r simple-search, eval = live_examples}
kinase_hits <- query_search("protein kinase")

head(kinase_hits, 10)
class(kinase_hits)
attr(kinase_hits, "return_type")
```

`query_search()` is the fastest way to ask a general question of the archive.
Here it performs a full-text search and returns entry identifiers. The returned
object is not just a plain character vector: it carries the class
`rPDBapi_query_ids`, which makes the contract explicit and helps downstream code
reason about what kind of object was returned.

# Workflow 2: Refine the Search with Structured Operators

```{r advanced-search, eval = live_examples}
kinase_entry_ids <- perform_search(
  search_operator = kinase_query,
  return_type = "ENTRY",
  request_options = search_controls,
  verbosity = FALSE
)

kinase_entry_ids
class(kinase_entry_ids)
```

`perform_search()` executes the operator-based query assembled earlier. This is
the function to use when you need precise control over attributes, logical
combinations, return types, or pagination. In structural bioinformatics, this
kind of targeted search is often more useful than full-text search alone,
because it lets you combine biological meaning with experimental constraints.

# Workflow 3: Retrieve Entry-Level Metadata

```{r entry-properties}
entry_properties <- list(
  rcsb_id = list(),
  struct = c("title"),
  struct_keywords = c("pdbx_keywords"),
  exptl = c("method"),
  rcsb_entry_info = c("molecular_weight", "resolution_combined"),
  rcsb_accession_info = c("initial_release_date")
)

entry_properties
```

This property list defines the fields we want from the GraphQL endpoint. It
captures both structural metadata and biologically meaningful annotations:
  structure title, keywords, experimental method, molecular weight, resolution,
and release date. By stating these fields explicitly, the workflow remains
transparent and easy to reproduce.

```{r schema-aware-properties}
head(list_rcsb_fields("ENTRY"), 10)
search_rcsb_fields("resolution", data_type = "ENTRY")

validate_properties(
  properties = entry_properties,
  data_type = "ENTRY",
  strict = TRUE
)

validate_properties(
  properties = list(
    rcsb_entry_info = c("resolution_combined", "unknown_subfield")
  ),
  data_type = "ENTRY",
  strict = FALSE
)
```

The schema-aware helpers are useful when building property lists iteratively.
`list_rcsb_fields()` exposes the package's built-in field registry,
`search_rcsb_fields()` narrows it to a topic of interest, and
`validate_properties()` checks that a property list matches the expected
data-type-specific structure. In strict mode, validation fails early; in
non-strict mode, it returns diagnostics that can be incorporated into an
interactive workflow or a package test.

```{r strict-validation-pattern, eval = FALSE}
old_opt <- options(rPDBapi.strict_property_validation = TRUE)
on.exit(options(old_opt), add = TRUE)

generate_json_query(
  ids = c("4HHB"),
  data_type = "ENTRY",
  properties = list(rcsb_entry_info = c("resolution_combined"))
)
```

This option-gated strict mode is useful when you want a pipeline to reject
unknown fields immediately. Because the option is opt-in, the package preserves
backward compatibility for existing code while still supporting stricter
validation in controlled workflows.

```{r entry-metadata, eval = live_examples}
kinase_metadata <- data_fetcher(
  id = kinase_entry_ids[1:5],
  data_type = "ENTRY",
  properties = entry_properties,
  return_as_dataframe = TRUE
)

kinase_metadata
```

`data_fetcher()` is the main high-level retrieval function for metadata. It
takes identifiers, the data level of interest, and a property list, then
returns either a validated nested response or a flattened data frame. For many
analysis tasks, returning a data frame is the most convenient choice because it
fits directly into standard R workflows for filtering, joining, and plotting.

# Workflow 4: Inspect the Raw API Payload and Convert It to Tidy Data

```{r raw-query, eval = live_examples}
kinase_json_query <- generate_json_query(
  ids = kinase_entry_ids[1:3],
  data_type = "ENTRY",
  properties = entry_properties
)

cat(kinase_json_query)
```

This chunk exposes the GraphQL query string that `rPDBapi` sends to the RCSB
data API. Seeing the generated query is helpful when you are debugging a field
name, comparing package output with the official schema, or teaching others how
the package maps R objects to API requests.

```{r raw-response, eval = live_examples}
kinase_raw <- fetch_data(
  json_query = kinase_json_query,
  data_type = "ENTRY",
  ids = kinase_entry_ids[1:3]
)

str(kinase_raw, max.level = 2)
```

`fetch_data()` returns a validated raw payload and tags it with the class
`rPDBapi_fetch_response`. This is useful when you want to inspect nested JSON
content before flattening it, preserve hierarchy for custom parsing, or verify
that a field is present before building a larger workflow around it.

```{r tidy-conversion, eval = live_examples}
kinase_tidy <- return_data_as_dataframe(
  response = kinase_raw,
  data_type = "ENTRY",
  ids = kinase_entry_ids[1:3]
)

kinase_tidy
```

`return_data_as_dataframe()` converts the nested response into a rectangular R
data structure. This transformation is central to reproducible bioinformatics:
once the results are tidy, they can be analyzed with `dplyr`, joined to other
annotations, summarized statistically, or passed to visualization packages.

# Workflow 4b: Batch Retrieval, Provenance, and Cache-Aware Access

High-throughput structural workflows rarely stop at one or two identifiers. In
screening, comparative analysis, or annotation projects, it is common to fetch
dozens or hundreds of records with the same property specification.

```{r batch-fetch, eval = live_examples}
cache_dir <- file.path(tempdir(), "rpdbapi-vignette-cache")

kinase_batch <- data_fetcher_batch(
  id = kinase_entry_ids[1:5],
  data_type = "ENTRY",
  properties = entry_properties,
  return_as_dataframe = TRUE,
  batch_size = 2,
  retry_attempts = 2,
  retry_backoff = 0,
  cache = TRUE,
  cache_dir = cache_dir,
  progress = FALSE,
  verbosity = FALSE
)

kinase_batch
attr(kinase_batch, "provenance")
cache_info(cache_dir)
```

`data_fetcher_batch()` scales the single-request `data_fetcher()` model to
larger identifier sets. It splits requests into batches, retries transient
failures, optionally stores batch results on disk, and attaches provenance to
the returned object. That provenance is important for reproducibility because it
records the retrieval mode, batch size, retry configuration, and cache usage.

```{r clear-cache, eval = live_examples}
clear_rpdbapi_cache(cache_dir)
cache_info(cache_dir)
```

This cache-management pattern is especially useful in iterative analysis.
Repeated metadata retrieval becomes faster when the same requests are reused,
while explicit cache inspection and cleanup keep the workflow transparent.

```{r batch-strategy, eval = FALSE}
# Use data_fetcher() when:
# - the ID set is small
# - you want the simplest request path
# - retry, cache, and provenance are unnecessary

# Use data_fetcher_batch() when:
# - the ID set is large
# - requests may need retries
# - repeated retrieval should reuse cached results
# - you want an explicit provenance record
```

In practice, `data_fetcher()` is usually sufficient for exploratory work.
`data_fetcher_batch()` becomes more useful as the workflow moves toward larger
or repeated retrieval, where retry behavior, caching, and provenance become
part of the analysis design rather than implementation detail.

```{r provenance-interpretation, eval = live_examples}
provenance_tbl <- dplyr::tibble(
  field = names(attr(kinase_batch, "provenance")),
  value = vapply(
    attr(kinase_batch, "provenance"),
    function(x) {
      if (is.list(x)) "<list>" else as.character(x)
    },
    character(1)
  )
)

provenance_tbl
```

Interpreting provenance explicitly is useful when results are produced in a
non-interactive workflow. The provenance record makes it clear how many batches
were used, whether caching was enabled, and how the retrieval was configured,
which makes the metadata table easier to audit later.

# Workflow 5: Retrieve Assembly-Level Data

Biological assemblies are often the correct unit of interpretation for
oligomeric proteins. A deposited asymmetric unit may not reflect the functional
quaternary structure, so assembly-level retrieval is important when studying
stoichiometry, symmetry, and interfaces.

```{r assembly-search, eval = live_examples}
kinase_assembly_ids <- perform_search(
  search_operator = kinase_query,
  return_type = "ASSEMBLY",
  request_options = RequestOptions(result_start_index = 0, num_results = 5),
  verbosity = FALSE
)

kinase_assembly_ids
```

This search requests assembly identifiers rather than entry identifiers. The
returned IDs encode both the entry and the assembly number, making them
appropriate inputs for assembly-level metadata retrieval. This is an important
distinction in structural biology because entry-level and assembly-level
questions are not interchangeable.

```{r assembly-metadata, eval = live_examples}
assembly_properties <- list(
  rcsb_id = list(),
  pdbx_struct_assembly = c("details", "method_details", "oligomeric_count"),
  rcsb_struct_symmetry = c("kind", "symbol")
)

kinase_assemblies <- data_fetcher(
  id = kinase_assembly_ids,
  data_type = "ASSEMBLY",
  properties = assembly_properties,
  return_as_dataframe = TRUE
)

kinase_assemblies
```

This chunk retrieves assembly descriptors and symmetry annotations. In
practice, these fields help answer questions about oligomeric state and
biological interpretation, such as whether a kinase structure is monomeric,
dimeric, or associated with a symmetric assembly.

```{r assembly-objects, eval = live_examples}
assembly_object <- as_rpdb_assembly(
  kinase_assemblies,
  metadata = list(query = "protein kinase assemblies")
)

assembly_object
dplyr::as_tibble(assembly_object)
summarize_assemblies(assembly_object)
```

The assembly object wrapper is useful when you want to retain lightweight
metadata alongside a table while still working with tibble-oriented tools.
`summarize_assemblies()` then provides a narrow helper for common assembly
questions, such as typical oligomeric count and the diversity of symmetry
labels in the retrieved result set.

# Workflow 5b: Identifier-Aware Retrieval Patterns

One practical source of bugs in structural workflows is mixing identifier
levels. A valid entry ID is not automatically a valid assembly or entity ID,
and the corresponding `data_type` must match the biological level of the
request.

```{r identifier-aware-retrieval}
dplyr::tibble(
  example_id = c("4HHB", "4HHB-1", "4HHB_1", "4HHB.A", "ATP"),
  inferred_type = infer_id_type(c("4HHB", "4HHB-1", "4HHB_1", "4HHB.A", "ATP"))
)

parse_rcsb_id("4HHB.A")
```

This is useful as a preflight step before retrieval. In larger workflows, a
small identifier check often saves time because it catches level mismatches
before the request reaches the API.

```{r identifier-aware-fetch, eval = FALSE}
# Entry-level retrieval
data_fetcher(
  id = build_entry_id("4HHB"),
  data_type = "ENTRY",
  properties = list(rcsb_id = list())
)

# Assembly-level retrieval
data_fetcher(
  id = build_assembly_id("4HHB", 1),
  data_type = "ASSEMBLY",
  properties = list(rcsb_id = list())
)

# Polymer-entity retrieval
data_fetcher(
  id = build_entity_id("4HHB", 1),
  data_type = "POLYMER_ENTITY",
  properties = list(rcsb_id = list())
)
```

Using the builder helpers makes the intended record level explicit in code.
That is especially helpful when identifiers are generated programmatically from
entry IDs and entity or assembly indices.

# Workflow 6: Retrieve Taxonomy and Chain-Level Biological Context

Many analyses need entity-level annotations rather than whole-entry metadata.
For example, taxonomy belongs naturally to polymer entities because different
entities within the same structure can come from different organisms or
constructs.

```{r polymer-search, eval = live_examples}
kinase_polymer_ids <- perform_search(
  search_operator = kinase_query,
  return_type = "POLYMER_ENTITY",
  request_options = RequestOptions(result_start_index = 0, num_results = 5),
  verbosity = FALSE
)

kinase_polymer_ids
```

Here the same biological query is projected onto polymer entities. This is
useful when you want annotations at the chain-definition level, such as source
organism, sequence grouping, or entity-specific descriptors.

```{r polymer-metadata, eval = live_examples}
polymer_properties <- list(
  rcsb_id = list(),
  rcsb_entity_source_organism = c("ncbi_taxonomy_id", "ncbi_scientific_name"),
  rcsb_cluster_membership = c("cluster_id", "identity")
)

kinase_polymer_metadata <- data_fetcher(
  id = kinase_polymer_ids,
  data_type = "POLYMER_ENTITY",
  properties = polymer_properties,
  return_as_dataframe = TRUE
)

kinase_polymer_metadata
```

This result provides organism-level context that is often essential in
comparative structural biology. For example, you might use these fields to
separate human kinase structures from bacterial homologs, or to identify
closely related entities before selecting representatives for downstream
modeling.

```{r taxonomy-extraction, eval = live_examples}
polymer_object <- as_rpdb_polymer_entity(
  kinase_polymer_metadata,
  metadata = list(query = "kinase polymer entities")
)

taxonomy_table <- extract_taxonomy_table(polymer_object)

taxonomy_table
taxonomy_table %>%
  count(ncbi_scientific_name, sort = TRUE)
```

`extract_taxonomy_table()` is intentionally narrow: it keeps only the fields
needed to represent source-organism assignments cleanly. This is useful when a
larger polymer-entity table contains many retrieval columns, but the immediate
analysis question is taxonomic composition or species-level redundancy.

# Workflow 7: Retrieve Detailed Entry Annotations

```{r entry-detail, eval = live_examples}
selected_entry <- kinase_entry_ids[[1]]
selected_info <- get_info(selected_entry)

entry_summary <- dplyr::tibble(
  rcsb_id = selected_entry,
  title = purrr::pluck(selected_info, "struct", "title", .default = NA_character_),
  keywords = purrr::pluck(selected_info, "struct_keywords", "pdbx_keywords", .default = NA_character_),
  method = purrr::pluck(selected_info, "exptl", 1, "method", .default = NA_character_),
  citation_title = purrr::pluck(selected_info, "rcsb_primary_citation", "title", .default = NA_character_),
  resolution = paste(
    purrr::pluck(selected_info, "rcsb_entry_info", "resolution_combined", .default = NA),
    collapse = "; "
  )
)

entry_summary
```

`get_info()` retrieves a full entry record as a nested list. This is useful
when you want a richer, less filtered representation than a GraphQL property
subset. In this example, we extract structure title, keywords, experimental
method, citation title, and resolution to build a compact summary of one kinase
entry. These fields are exactly the kinds of annotations structural biologists
inspect when deciding whether a structure is suitable for biological
interpretation or downstream modeling.

```{r literature-links, eval = live_examples}
kinase_papers <- find_papers("protein kinase", max_results = 3)
kinase_keywords <- find_results("protein kinase", field = "struct_keywords")

kinase_papers
utils::head(kinase_keywords, 3)
```

These helper functions show how `rPDBapi` can bridge structures and biological
interpretation. `find_papers()` provides publication titles associated with
matching entries, while `find_results()` can retrieve selected metadata fields
across search results. Together, they support a workflow in which structures,
annotations, and literature references are linked inside the same R session.

# Workflow 8: Download Coordinates and Inspect Atomic Data

```{r coordinates, eval = live_examples}
kinase_structure <- get_pdb_file(
  pdb_id = selected_entry,
  filetype = "cif",
  verbosity = FALSE
)

coordinate_matrix <- matrix(kinase_structure$xyz, ncol = 3, byrow = TRUE)
coordinate_df <- data.frame(
  x = coordinate_matrix[, 1],
  y = coordinate_matrix[, 2],
  z = coordinate_matrix[, 3]
)

calpha_atoms <- cbind(
  kinase_structure$atom[kinase_structure$calpha, c("chain", "resno", "resid")],
  coordinate_df[kinase_structure$calpha, , drop = FALSE]
)

head(calpha_atoms, 10)
```

`get_pdb_file()` downloads and parses the structure file into an object that
contains atomic records and coordinates. This is the transition point from
metadata analysis to coordinate analysis. The example extracts C-alpha atoms,
which are commonly used in structural alignment, geometry summaries, coarse
distance analyses, and quick visual inspection of protein backbones.

```{r calpha-helper, eval = live_examples}
calpha_atoms <- extract_calpha_coordinates(kinase_structure)

head(calpha_atoms, 10)
```

`extract_calpha_coordinates()` packages a common structural-bioinformatics step
into a reusable helper. The result is immediately usable for plotting,
distance-based summaries, or chain-level coordinate analyses without manually
reconstructing the atom/coordinate join each time.

```{r fasta, eval = live_examples}
kinase_sequences <- get_fasta_from_rcsb_entry(selected_entry, verbosity = FALSE)

length(kinase_sequences)
utils::head(nchar(unlist(kinase_sequences)))
```

The FASTA workflow complements coordinate retrieval by exposing the underlying
macromolecular sequences. Having both sequence and structure available in the
same environment is useful for tasks such as domain boundary checks, sequence
length summaries, and linking structural hits to sequence-based pipelines.

```{r structure-sequence-join, eval = live_examples}
chain_sequence_summary <- join_structure_sequence(
  kinase_structure,
  kinase_sequences
)

chain_sequence_summary
```

This helper joins sequence-level and coordinate-level summaries at the chain
level. In practice, it provides a quick diagnostic for whether the downloaded
structure and the FASTA content align as expected, and it creates a compact
table that can be extended with additional chain annotations.

# Workflow 8b: Working with the Rich Object Model

The typed object wrappers are intentionally lightweight. They do not replace
the underlying data; instead, they add a stable class layer for printing,
conversion, and helper dispatch.

```{r object-model-local}
entry_demo <- as_rpdb_entry(
  data.frame(
    rcsb_id = c("4HHB", "1CRN"),
    method = c("X-RAY DIFFRACTION", "SOLUTION NMR"),
    resolution_combined = c("1.74", NA),
    stringsAsFactors = FALSE
  ),
  metadata = list(example = "local object demo")
)

entry_demo
dplyr::as_tibble(entry_demo)
summarize_entries(entry_demo)
entry_demo$metadata
```

This pattern is useful when a workflow needs to preserve both data and context.
For example, metadata can record which query produced a table or which
processing choices were applied. The `as_tibble()` methods then let the object
drop back into a standard tidyverse pipeline without extra conversion code.

```{r structure-object-local}
structure_demo <- as_rpdb_structure(
  list(
    atom = data.frame(
      chain = c("A", "A"),
      resno = c(1L, 2L),
      resid = c("GLY", "ALA"),
      stringsAsFactors = FALSE
    ),
    xyz = c(1, 2, 3, 4, 5, 6),
    calpha = c(TRUE, FALSE)
  ),
  metadata = list(source = "illustration")
)

structure_demo
dplyr::as_tibble(structure_demo)
```

The structure wrapper is especially useful when one analysis session contains
multiple parsed structures and derived tables. The class layer makes those
objects easier to distinguish and easier to handle consistently.

# Workflow 9: Downstream Analysis in R

```{r downstream-analysis, eval = live_examples}
entry_object <- as_rpdb_entry(
  kinase_metadata,
  metadata = list(query = "protein kinase entry metadata")
)

summarize_entries(entry_object)

kinase_summary <- dplyr::as_tibble(entry_object) %>%
  mutate(
    molecular_weight = as.numeric(molecular_weight),
    resolution_combined = as.numeric(resolution_combined),
    initial_release_date = as.Date(initial_release_date)
  ) %>%
  arrange(resolution_combined) %>%
  select(
    rcsb_id,
    title,
    pdbx_keywords,
    method,
    molecular_weight,
    resolution_combined,
    initial_release_date
  )

kinase_summary

kinase_summary %>%
  summarise(
    n_structures = n(),
    median_molecular_weight = median(molecular_weight, na.rm = TRUE),
    best_resolution = min(resolution_combined, na.rm = TRUE)
  )
```

Once the metadata are in a data frame, standard R analysis becomes immediate.
This chunk ranks the retrieved kinase structures by resolution and computes a
few simple summaries. Although the analysis is straightforward, it illustrates
the main advantage of using `rPDBapi`: PDB metadata become ordinary tabular R
data that can be manipulated with the same tools used elsewhere in
bioinformatics.

```{r taxonomy-summary, eval = live_examples}
kinase_polymer_metadata %>%
  count(ncbi_scientific_name, sort = TRUE)
```

This example summarizes the source organisms represented in the polymer-entity
results. A table like this is often the first step in identifying redundancy,
sampling bias, or opportunities to compare orthologous structures across
species.

# Workflow 10: Optional Visualization with r3dmol

```{r r3dmol-view, eval = live_examples && have_r3dmol}
saved_structure <- get_pdb_file(
  pdb_id = selected_entry,
  filetype = "pdb",
  save = TRUE,
  path = tempdir(),
  verbosity = FALSE
)

r3dmol::r3dmol() %>%
  r3dmol::m_add_model(data = saved_structure$path, format = "pdb") %>%
  r3dmol::m_set_style(style = r3dmol::m_style_cartoon(color = "spectrum")) %>%
  r3dmol::m_zoom_to()
```

This optional chunk demonstrates how `rPDBapi` fits into broader R
visualization workflows. The package itself focuses on data access and parsing,
while a tool such as `r3dmol` can be used to render the retrieved structure in
3D. That separation of responsibilities is useful because it keeps data access,
analysis, and visualization composable.

# Advanced Search Modalities

The previous sections used text-based and attribute-based search. `rPDBapi`
also supports sequence, motif, structure, and chemical searches. These are
especially important in structural bioinformatics, where the biological question
is often not "Which entry mentions a keyword?" but "Which structures resemble
this sequence, motif, fold, or ligand chemistry?"

## Sequence Search

```{r sequence-operator}
kinase_motif_sequence <- "VAIKTLKPGTMSPEAFLQEAQVMKKLRHEKLVQLYAVV"

sequence_operator <- SequenceOperator(
  sequence = kinase_motif_sequence,
  sequence_type = "PROTEIN",
  evalue_cutoff = 10,
  identity_cutoff = 0.7
)

sequence_operator
autoresolve_sequence_type("ATGCGTACGTAGC")
autoresolve_sequence_type("AUGCGUACGUAGC")
```

`SequenceOperator()` constructs a sequence-similarity search request. This is
useful when you have a sequence of interest, such as a kinase domain fragment,
and want to find structurally characterized homologs in the PDB. The companion
function `autoresolve_sequence_type()` infers whether a sequence is DNA, RNA,
or protein, which is helpful for interactive workflows and programmatic
pipelines that ingest sequences from external sources.

```{r sequence-search, eval = live_examples}
sequence_hits <- perform_search(
  search_operator = sequence_operator,
  return_type = "POLYMER_ENTITY",
  request_options = RequestOptions(result_start_index = 0, num_results = 5),
  verbosity = FALSE
)

sequence_hits
```

This query asks the RCSB search service for polymer entities similar to the
input sequence. The polymer-entity return type is appropriate here because
sequence similarity is defined at the entity level rather than at the whole
entry level.

## Sequence Motif Search

```{r seqmotif-operator}
prosite_like_motif <- SeqMotifOperator(
  pattern = "[LIV]-x-K-[GST]",
  sequence_type = "PROTEIN",
  pattern_type = "REGEX"
)

prosite_like_motif
```

`SeqMotifOperator()` targets local sequence patterns rather than whole-sequence
similarity. This is useful for catalytic signatures, binding motifs, or
short conserved regions that occur across otherwise diverse proteins.

```{r seqmotif-search, eval = live_examples}
motif_hits <- perform_search(
  search_operator = prosite_like_motif,
  return_type = "POLYMER_ENTITY",
  request_options = RequestOptions(result_start_index = 0, num_results = 5),
  verbosity = FALSE
)

motif_hits
```

Motif search is especially helpful when a biological question depends on a
functional local pattern rather than on full-length homology. In kinase-related
work, motifs can help you locate catalytic or regulatory sequence signatures
across structural entries.

## Structure Similarity Search

```{r structure-operator}
structure_operator <- StructureOperator(
  pdb_entry_id = "4HHB",
  assembly_id = 1,
  search_mode = "RELAXED_SHAPE_MATCH"
)

structure_operator
infer_search_service(structure_operator)
```

`StructureOperator()` creates a shape-based structure search using an existing
PDB structure as the template. This is useful when you want to identify
structures with related global geometry, even if sequence identity is limited.
`infer_search_service()` confirms that this operator is routed to the structure
search backend rather than to text or full-text search.

```{r structure-search, eval = live_examples}
structure_hits <- perform_search(
  search_operator = structure_operator,
  return_type = "ASSEMBLY",
  request_options = RequestOptions(result_start_index = 0, num_results = 5),
  verbosity = FALSE
)

structure_hits
```

Returning assemblies is often sensible for shape-based search because biological
function frequently depends on the quaternary arrangement of chains rather than
only on the deposited asymmetric unit.

## Chemical Search

```{r chemical-operator}
atp_like_operator <- ChemicalOperator(
  descriptor = "O=P(O)(O)OP(=O)(O)OP(=O)(O)O",
  matching_criterion = "fingerprint-similarity"
)

atp_like_operator
infer_search_service(atp_like_operator)
```

`ChemicalOperator()` supports ligand-centric workflows by searching the archive
with a SMILES or InChI descriptor. This is useful when you want to find
structures containing chemically similar ligands, cofactors, or fragments.

```{r chemical-search, eval = live_examples}
chemical_hits <- perform_search(
  search_operator = atp_like_operator,
  return_type = "CHEMICAL_COMPONENT",
  request_options = RequestOptions(result_start_index = 0, num_results = 5),
  verbosity = FALSE
)

chemical_hits
```

Here the return type is `CHEMICAL_COMPONENT`, which maps to molecular
definitions in the RCSB search API. This lets the search focus on ligand
identities rather than on whole macromolecular entries.

# Complete Search Operator Reference

The operator-based search grammar is one of the most important features of the
package. The examples below summarize the text-oriented operators that can be
combined in grouped queries.

```{r operator-reference}
exact_resolution <- ExactMatchOperator(
  attribute = "exptl.method",
  value = "X-RAY DIFFRACTION"
)

organism_inclusion <- InOperator(
  attribute = "rcsb_entity_source_organism.taxonomy_lineage.name",
  value = c("Homo sapiens", "Mus musculus")
)

title_words <- ContainsWordsOperator(
  attribute = "struct.title",
  value = "protein kinase"
)

title_phrase <- ContainsPhraseOperator(
  attribute = "struct.title",
  value = "protein kinase"
)

resolution_cutoff <- ComparisonOperator(
  attribute = "rcsb_entry_info.resolution_combined",
  value = 2.0,
  comparison_type = "LESS"
)

resolution_window <- RangeOperator(
  attribute = "rcsb_entry_info.resolution_combined",
  from_value = 1.0,
  to_value = 2.5
)

doi_exists <- ExistsOperator("rcsb_primary_citation.pdbx_database_id_doi")

list(
  exact_resolution = exact_resolution,
  organism_inclusion = organism_inclusion,
  title_words = title_words,
  title_phrase = title_phrase,
  resolution_cutoff = resolution_cutoff,
  resolution_window = resolution_window,
  doi_exists = doi_exists
)
```

These operator constructors do not perform a search by themselves. Instead,
they make query intent explicit and composable. That matters when analyses need
to be read, reviewed, and reproduced months later.

```{r querynode-scoredresult}
operator_node <- QueryNode(title_words)

composite_query <- QueryGroup(
  queries = list(title_words, resolution_window, doi_exists),
  logical_operator = "AND"
)

scored_example <- ScoredResult(entity_id = "4HHB", score = 0.98)

operator_node
composite_query
scored_example
```

`QueryNode()` and `QueryGroup()` are the glue that turn independent operator
objects into a full search graph. `ScoredResult()` is a small utility that
represents the shape of a scored hit and is useful for result handling or for
teaching the output model used by structure-search APIs.

```{r scored-search-results, eval = live_examples}
scored_structure_hits <- perform_search(
  search_operator = structure_operator,
  return_type = "ASSEMBLY",
  request_options = RequestOptions(result_start_index = 0, num_results = 3),
  return_with_scores = TRUE,
  verbosity = FALSE
)

scored_structure_hits
class(scored_structure_hits)
```

This example shows the difference between returning plain identifiers and
returning scored hits. In similarity-oriented workflows, the score itself can
be analytically useful because it helps rank follow-up candidates before any
metadata are fetched. A common pattern is to inspect scored results first,
choose a cutoff, and only then retrieve metadata for the retained identifiers.

```{r query-composition-strategy, eval = FALSE}
# Pattern: build small reusable operators first
title_filter <- ContainsPhraseOperator("struct.title", "protein kinase")
resolution_filter <- ComparisonOperator(
  "rcsb_entry_info.resolution_combined",
  2.5,
  "LESS_OR_EQUAL"
)

# Wrap or combine them only when the biological question is clear
query_graph <- QueryGroup(
  queries = list(
    QueryNode(title_filter),
    QueryNode(resolution_filter)
  ),
  logical_operator = "AND"
)
```

This is the main reason to treat operator construction as a separate step. A
query graph can be assembled gradually, reviewed independently of the network
request, and reused across multiple searches or result types.

# Query Search Variants and Scan Parameters

`query_search()` is intentionally simpler than `perform_search()`, but it still
supports several specialized query types as well as low-level request overrides.

```{r query-search-variants, eval = FALSE}
# PubMed-linked structures
query_search(search_term = 27499440, query_type = "PubmedIdQuery")

# Organism/taxonomy search
query_search(search_term = "9606", query_type = "TreeEntityQuery")

# Experimental method search
query_search(search_term = "X-RAY DIFFRACTION", query_type = "ExpTypeQuery")

# Author search
query_search(search_term = "Kuriyan, J.", query_type = "AdvancedAuthorQuery")

# UniProt-linked entries
query_search(search_term = "P31749", query_type = "uniprot")

# PFAM-linked entries
query_search(search_term = "PF00069", query_type = "pfam")
```

These convenience modes are useful when the search criterion maps directly to a
common biological identifier or curation field. They are less flexible than a
fully operator-based query, but faster to write for routine tasks.

```{r scan-params-example}
custom_scan_params <- list(
  request_options = list(
    paginate = list(start = 0, rows = 5),
    return_all_hits = FALSE
  )
)

custom_scan_params
```

`scan_params` lets you override the request body that `query_search()` sends to
the search API. This is useful when you want lightweight access to custom
pagination or request options without switching fully to `perform_search()`.

```{r query-search-scan-params, eval = live_examples}
limited_kinase_hits <- query_search(
  search_term = "protein kinase",
  scan_params = custom_scan_params
)

limited_kinase_hits
```

This example illustrates the practical use of `scan_params`: constrain the
result set while still using the simpler query helper.

# Complete Metadata Retrieval Surface

The `data_fetcher()` interface supports more than entry and polymer-entity data.
The main supported data types are:

- `ENTRY`
- `ASSEMBLY`
- `POLYMER_ENTITY`
- `BRANCHED_ENTITY`
- `NONPOLYMER_ENTITY`
- `POLYMER_ENTITY_INSTANCE`
- `BRANCHED_ENTITY_INSTANCE`
- `NONPOLYMER_ENTITY_INSTANCE`
- `CHEMICAL_COMPONENT`

This breadth matters because structural records are hierarchical. Different
questions belong to different levels: entry-level methods, assembly-level
symmetry, entity-level taxonomy, instance-level chain annotations, and
component-level ligand chemistry.

## Building Property Lists Incrementally

```{r add-property}
base_properties <- list(
  rcsb_entry_info = c("resolution_combined"),
  exptl = c("method")
)

extended_properties <- add_property(list(
  rcsb_entry_info = c("molecular_weight", "resolution_combined"),
  struct = c("title")
))

base_properties
extended_properties
```

`add_property()` helps construct or merge property specifications without
duplicating subfields. This is especially useful in interactive analyses, where
you may start with a minimal query and then progressively request additional
annotations.

```{r property-design-pattern}
property_workflow <- add_property(list(
  rcsb_id = list(),
  struct = c("title"),
  rcsb_entry_info = c("resolution_combined")
))

property_workflow <- add_property(list(
  rcsb_entry_info = c("molecular_weight", "resolution_combined"),
  exptl = c("method")
))

property_workflow
validate_properties(property_workflow, data_type = "ENTRY", strict = FALSE)
```

This pattern is useful because GraphQL property lists tend to grow as an
analysis becomes more specific. Building them incrementally makes it easier to
keep a compact initial query, add only the fields that become necessary, and
check that the evolving specification still matches the expected schema.

## Non-polymer and Chemical Component Data

```{r ligand-component-properties}
ligand_properties <- list(
  rcsb_id = list(),
  chem_comp = c("id", "name", "formula", "formula_weight", "type"),
  rcsb_chem_comp_info = c("initial_release_date")
)

ligand_properties
```

This property specification targets chemical components rather than whole
structures. That distinction is important when the biological focus is on
ligands, cofactors, inhibitors, or bound metabolites.

```{r chemical-component-fetch, eval = live_examples}
chemical_component_df <- data_fetcher(
  id = head(chemical_hits, 3),
  data_type = "CHEMICAL_COMPONENT",
  properties = ligand_properties,
  return_as_dataframe = TRUE
)

chemical_component_df
```

The resulting table can be used to compare ligand formulas, molecular weights,
and release histories. This is often useful in medicinal chemistry and
structure-based design workflows.

```{r ligand-object-helper, eval = live_examples}
ligand_object <- as_rpdb_chemical_component(
  chemical_component_df,
  metadata = list(query = "ATP-like chemical components")
)

extract_ligand_table(ligand_object)
```

`extract_ligand_table()` keeps the most analysis-relevant chemical-component
columns in a compact form. That is useful when ligand retrieval is only one
part of a broader workflow and you want a small, stable table for downstream
joins or ranking.

```{r describe-chemical, eval = live_examples}
atp_description <- describe_chemical("ATP")

dplyr::tibble(
  chem_id = "ATP",
  name = purrr::pluck(atp_description, "chem_comp", "name", .default = NA_character_),
  formula = purrr::pluck(atp_description, "chem_comp", "formula", .default = NA_character_),
  formula_weight = purrr::pluck(atp_description, "chem_comp", "formula_weight", .default = NA),
  smiles = purrr::pluck(atp_description, "rcsb_chem_comp_descriptor", "smiles", .default = NA_character_)
)
```

`describe_chemical()` provides a direct route to detailed ligand information for
a single chemical component. It complements `data_fetcher()` by supporting a
focused, ligand-centric lookup.

## Instance-Level Retrieval

```{r instance-level-examples, eval = FALSE}
# Polymer chain instances
data_fetcher(
  id = c("4HHB.A"),
  data_type = "POLYMER_ENTITY_INSTANCE",
  properties = list(rcsb_id = list())
)

# Branched entity instances
data_fetcher(
  id = c("1ABC_2"),
  data_type = "BRANCHED_ENTITY_INSTANCE",
  properties = list(rcsb_id = list())
)

# Non-polymer entity instances
data_fetcher(
  id = c("3PQR_5"),
  data_type = "NONPOLYMER_ENTITY_INSTANCE",
  properties = list(rcsb_id = list())
)
```

Instance-level retrieval is relevant when chain-level or site-specific
annotations matter. The exact identifier format depends on the corresponding
RCSB data type and record level.

# Low-Level API Access and Parsing Helpers

The package exposes lower-level functions for users who need full control over
HTTP requests, URLs, or response parsing.

```{r low-level-url}
entry_url <- get_pdb_api_url("core/entry/", "4HHB")
chem_url <- get_pdb_api_url("core/chemcomp/", "ATP")

entry_url
chem_url
```

`get_pdb_api_url()` constructs endpoint-specific URLs. This is a small utility,
but it makes low-level workflows clearer and reduces hard-coded URL strings in
custom scripts.

```{r low-level-lifecycle, eval = FALSE}
# Manual request lifecycle
url <- get_pdb_api_url("core/entry/", "4HHB")
response <- send_api_request(url, verbosity = FALSE)
handle_api_errors(response, url)
payload <- parse_response(response, format = "json")
```

This low-level lifecycle is useful when you are developing a new helper,
debugging an endpoint transition, or comparing package behavior with the raw
RCSB API. It also makes clear where URL construction, HTTP transport, status
checking, and parsing are separated inside the package.

```{r low-level-http, eval = live_examples}
entry_response <- send_api_request(entry_url, verbosity = FALSE)
handle_api_errors(entry_response, entry_url)
entry_payload <- parse_response(entry_response, format = "json")

names(entry_payload)[1:5]
```

These functions expose the package's low-level request stack. `send_api_request()`
handles the HTTP request, `handle_api_errors()` checks the returned status, and
`parse_response()` converts the body into an R object. This layer is useful
when you need to debug endpoint behavior or prototype a new helper around the
RCSB REST API.

```{r graphql-low-level, eval = live_examples}
mini_graphql <- generate_json_query(
  ids = kinase_entry_ids[1:2],
  data_type = "ENTRY",
  properties = list(rcsb_id = list(), struct = c("title"))
)

mini_graphql_response <- search_graphql(mini_graphql)

str(mini_graphql_response, max.level = 2)
```

`search_graphql()` is the low-level GraphQL entry point. It is useful when you
want to inspect the raw content returned by the RCSB GraphQL service before it
is normalized by `fetch_data()` or flattened by `return_data_as_dataframe()`.

# Return Contracts and Error Handling

One of the notable features of the current package version is that core
functions return typed objects. This improves programmatic safety because code
can distinguish search identifiers, scored results, raw responses, and flattened
data frames.

```{r contracts-live, eval = live_examples}
list(
  query_search_class = class(query_search("kinase")),
  perform_search_class = class(
    perform_search(DefaultOperator("kinase"), verbosity = FALSE)
  ),
  perform_search_scores_class = class(
    perform_search(
      DefaultOperator("kinase"),
      return_with_scores = TRUE,
      verbosity = FALSE
    )
  )
)
```

The classes shown above make return semantics explicit. In a larger analysis
pipeline, this reduces ambiguity and makes it easier to validate assumptions at
each stage of data retrieval.

```{r fetch-contracts, eval = live_examples}
raw_entry_response <- data_fetcher(
  id = kinase_entry_ids[1:2],
  data_type = "ENTRY",
  properties = list(rcsb_id = list()),
  return_as_dataframe = FALSE
)

tidy_entry_response <- data_fetcher(
  id = kinase_entry_ids[1:2],
  data_type = "ENTRY",
  properties = list(rcsb_id = list()),
  return_as_dataframe = TRUE
)

class(raw_entry_response)
class(tidy_entry_response)
```

This example emphasizes the dual output model of `data_fetcher()`: retain the
nested payload when structure matters, or request a data frame when analysis
and integration matter more.

```{r object-contracts, eval = live_examples}
list(
  entry_object_class = class(as_rpdb_entry(kinase_metadata)),
  assembly_object_class = class(as_rpdb_assembly(kinase_assemblies)),
  polymer_object_class = class(as_rpdb_polymer_entity(kinase_polymer_metadata)),
  structure_object_class = class(as_rpdb_structure(kinase_structure)),
  batch_provenance_names = names(attr(kinase_batch, "provenance"))
)
```

These richer object wrappers are deliberately lightweight. They preserve the
underlying data while attaching a semantically meaningful class, which makes it
easier to define helper methods and to branch on object type in larger
structural analysis pipelines.

```{r object-methods-local}
local_entry_object <- as_rpdb_entry(
  data.frame(
    rcsb_id = "4HHB",
    method = "X-RAY DIFFRACTION",
    resolution_combined = "1.74",
    stringsAsFactors = FALSE
  ),
  metadata = list(source = "local method demo")
)

print(local_entry_object)
dplyr::as_tibble(local_entry_object)
```

This example makes the object behavior explicit. The custom print method gives a
 concise summary of the wrapped object, while the `as_tibble()` method provides
an immediate path back to a standard tabular workflow. That combination is the
main point of the object model: preserve semantic type information without
making downstream manipulation cumbersome.

```{r defensive-patterns}
invalid_property_result <- tryCatch(
  validate_properties(
    properties = list(unknown_field = c("x")),
    data_type = "ENTRY",
    strict = TRUE
  ),
  rPDBapi_error_invalid_input = function(e) e
)

invalid_fetch_result <- tryCatch(
  data_fetcher(
    id = character(0),
    data_type = "ENTRY",
    properties = list(rcsb_id = list())
  ),
  rPDBapi_error_invalid_input = function(e) e
)

list(
  invalid_property_class = class(invalid_property_result),
  invalid_property_message = conditionMessage(invalid_property_result),
  invalid_fetch_class = class(invalid_fetch_result),
  invalid_fetch_message = conditionMessage(invalid_fetch_result)
)
```

These examples show how typed errors support defensive programming. Instead of
matching raw error text, a calling workflow can branch on the condition class
and decide whether to stop, retry, skip a record, or log the problem for later
review. That is particularly valuable when `rPDBapi` is used inside larger
automated pipelines.

# Appendix A: Export-by-Export Reference

The table below maps every exported function to its primary role in the package.
This is not a replacement for the individual help pages, but it does make the
full surface area of the package explicit inside a single tutorial document.

```{r export-reference, results = "asis"}
export_reference <- data.frame(
  Function = c(
    "query_search", "perform_search", "DefaultOperator", "ExactMatchOperator",
    "InOperator", "ContainsWordsOperator", "ContainsPhraseOperator",
    "ComparisonOperator", "RangeOperator", "ExistsOperator",
    "SequenceOperator", "autoresolve_sequence_type", "SeqMotifOperator",
    "StructureOperator", "ChemicalOperator", "QueryNode", "QueryGroup",
    "RequestOptions", "ScoredResult", "infer_search_service",
    "infer_id_type", "parse_rcsb_id", "build_entry_id", "build_assembly_id",
    "build_entity_id", "build_instance_id", "add_property",
    "list_rcsb_fields", "search_rcsb_fields", "validate_properties",
    "generate_json_query", "search_graphql", "fetch_data",
    "return_data_as_dataframe", "data_fetcher", "data_fetcher_batch",
    "cache_info", "clear_rpdbapi_cache", "get_info", "find_results",
    "find_papers", "describe_chemical", "get_fasta_from_rcsb_entry",
    "get_pdb_file", "get_pdb_api_url", "send_api_request",
    "handle_api_errors", "parse_response", "as_rpdb_entry",
    "as_rpdb_assembly", "as_rpdb_polymer_entity",
    "as_rpdb_chemical_component", "as_rpdb_structure",
    "summarize_entries", "summarize_assemblies",
    "extract_taxonomy_table", "extract_ligand_table",
    "extract_calpha_coordinates", "join_structure_sequence"
  ),
  Role = c(
    "High-level convenience search helper",
    "Operator-based search engine",
    "Full-text search operator",
    "Exact attribute match operator",
    "Set-membership operator",
    "Word containment operator",
    "Phrase containment operator",
    "Numeric/date comparison operator",
    "Range filter operator",
    "Attribute existence operator",
    "Sequence similarity search operator",
    "Automatic DNA/RNA/protein detection",
    "Sequence motif search operator",
    "Structure similarity search operator",
    "Chemical descriptor search operator",
    "Wrap one operator as a query node",
    "Combine nodes with AND/OR logic",
    "Pagination and sorting controls",
    "Represent a scored hit",
    "Infer backend service from operator",
    "Infer identifier level from an ID string",
    "Parse an identifier into structured components",
    "Normalize or build entry identifiers",
    "Build assembly identifiers",
    "Build entity identifiers",
    "Build instance or chain identifiers",
    "Merge/extend GraphQL property lists",
    "List known retrievable fields by data type",
    "Search the built-in field registry",
    "Validate a property list against the field registry",
    "Build a GraphQL query string",
    "Low-level GraphQL request helper",
    "Normalize validated GraphQL payloads",
    "Flatten nested payloads into data frames",
    "High-level metadata fetcher",
    "Batch metadata fetcher with retry and provenance",
    "Inspect batch-cache contents",
    "Clear on-disk cache entries",
    "Retrieve full entry metadata",
    "Extract one field across search hits",
    "Extract primary citation titles",
    "Retrieve ligand/chemical-component details",
    "Retrieve FASTA sequences",
    "Download and parse structure files",
    "Build REST endpoint URLs",
    "Send low-level GET/POST requests",
    "Check HTTP status and stop on error",
    "Parse JSON or text responses",
    "Wrap entry data in a typed object",
    "Wrap assembly data in a typed object",
    "Wrap polymer-entity data in a typed object",
    "Wrap chemical-component data in a typed object",
    "Wrap structure data in a typed object",
    "Summarize entry-level metadata",
    "Summarize assembly-level metadata",
    "Extract taxonomy-focused columns",
    "Extract ligand-focused columns",
    "Extract C-alpha coordinates",
    "Join sequence summaries to chain coordinates"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(export_reference, align = c("l", "l"))
```

This table is intended as a package navigation aid. It makes it easier to
identify whether a task belongs to searching, retrieval, parsing, or lower-level
API control before you start writing a workflow.

# Appendix B: Minimal Example Pattern for Every Export

The next block gives a compact usage sketch for every exported function. These
examples are deliberately short and grouped by role so that users can quickly
find a starting pattern.

```{r every-export-pattern, eval = FALSE}
# Search helpers
query_search("kinase")
perform_search(DefaultOperator("kinase"), verbosity = FALSE)

# Text and attribute operators
DefaultOperator("kinase")
ExactMatchOperator("exptl.method", "X-RAY DIFFRACTION")
InOperator("rcsb_entity_source_organism.taxonomy_lineage.name", c("Homo sapiens", "Mus musculus"))
ContainsWordsOperator("struct.title", "protein kinase")
ContainsPhraseOperator("struct.title", "protein kinase")
ComparisonOperator("rcsb_entry_info.resolution_combined", 2.0, "LESS")
RangeOperator("rcsb_entry_info.resolution_combined", 1.0, 2.5)
ExistsOperator("rcsb_primary_citation.pdbx_database_id_doi")

# Specialized operators
SequenceOperator("MVLSPADKTNVKAAW", sequence_type = "PROTEIN")
autoresolve_sequence_type("ATGCGTACGTAGC")
SeqMotifOperator("[LIV]-x-K-[GST]", "PROTEIN", "REGEX")
StructureOperator("4HHB", assembly_id = 1, search_mode = "RELAXED_SHAPE_MATCH")
ChemicalOperator("C1=CC=CC=C1", matching_criterion = "graph-strict")

# Query composition
QueryNode(DefaultOperator("kinase"))
QueryGroup(list(DefaultOperator("kinase"), ExistsOperator("rcsb_primary_citation.title")), "AND")
RequestOptions(result_start_index = 0, num_results = 10)
ScoredResult("4HHB", 0.98)
infer_search_service(StructureOperator("4HHB"))
infer_id_type(c("4HHB", "4HHB-1", "4HHB_1", "4HHB.A", "ATP"))
parse_rcsb_id("4HHB-1")
build_entry_id("4HHB")
build_assembly_id("4HHB", 1)
build_entity_id("4HHB", 1)
build_instance_id("4HHB", "A")

# Metadata helpers
add_property(list(rcsb_entry_info = c("resolution_combined")))
list_rcsb_fields("ENTRY")
search_rcsb_fields("resolution", data_type = "ENTRY")
validate_properties(
  list(rcsb_id = list(), rcsb_entry_info = c("resolution_combined")),
  data_type = "ENTRY",
  strict = TRUE
)
generate_json_query(c("4HHB"), "ENTRY", list(rcsb_id = list(), struct = c("title")))
search_graphql(generate_json_query(c("4HHB"), "ENTRY", list(rcsb_id = list())))
fetch_data(generate_json_query(c("4HHB"), "ENTRY", list(rcsb_id = list())), "ENTRY", "4HHB")
return_data_as_dataframe(
  fetch_data(generate_json_query(c("4HHB"), "ENTRY", list(rcsb_id = list())), "ENTRY", "4HHB"),
  "ENTRY",
  "4HHB"
)
data_fetcher("4HHB", "ENTRY", list(rcsb_id = list(), struct = c("title")))
data_fetcher_batch(
  c("4HHB", "1CRN"),
  "ENTRY",
  list(rcsb_id = list(), struct = c("title")),
  batch_size = 1,
  cache = FALSE
)
cache_info()
clear_rpdbapi_cache()
get_info("4HHB")
find_results("kinase", field = "struct_keywords")
find_papers("kinase", max_results = 3)
describe_chemical("ATP")
get_fasta_from_rcsb_entry("4HHB")

# Files and low-level HTTP
get_pdb_file("4HHB", filetype = "cif", verbosity = FALSE)
get_pdb_api_url("core/entry/", "4HHB")
resp <- send_api_request(get_pdb_api_url("core/entry/", "4HHB"), verbosity = FALSE)
handle_api_errors(resp, get_pdb_api_url("core/entry/", "4HHB"))
parse_response(resp, format = "json")

# Object wrappers and analysis helpers
as_rpdb_entry(data.frame(rcsb_id = "4HHB"))
as_rpdb_assembly(data.frame(rcsb_id = "4HHB-1"))
as_rpdb_polymer_entity(data.frame(rcsb_id = "4HHB_1"))
as_rpdb_chemical_component(data.frame(rcsb_id = "ATP"))
as_rpdb_structure(get_pdb_file("4HHB", filetype = "cif", verbosity = FALSE))
summarize_entries(data.frame(method = "X-RAY DIFFRACTION", resolution_combined = "1.8"))
summarize_assemblies(data.frame(oligomeric_count = "2", symbol = "C2"))
extract_taxonomy_table(data.frame(rcsb_id = "4HHB_1", ncbi_taxonomy_id = "9606"))
extract_ligand_table(data.frame(rcsb_id = "ATP", formula_weight = "507.18"))
extract_calpha_coordinates(get_pdb_file("4HHB", filetype = "cif", verbosity = FALSE))
join_structure_sequence(
  get_pdb_file("4HHB", filetype = "cif", verbosity = FALSE),
  get_fasta_from_rcsb_entry("4HHB")
)
```

This appendix is intentionally compact. Its purpose is not to replace the
narrative examples above, but to ensure that every exported function has an
immediately visible calling pattern in the vignette.

# Appendix C: Supported Identifier Levels and Typical Formats

One source of confusion in the RCSB ecosystem is that different endpoints expect
different identifier types. The table below summarizes the levels supported by
`data_fetcher()` and the search return types used in `perform_search()`.

```{r id-format-table, results = "asis"}
id_reference <- data.frame(
  Data_or_Return_Type = c(
    "ENTRY", "ASSEMBLY", "POLYMER_ENTITY", "BRANCHED_ENTITY",
    "NONPOLYMER_ENTITY", "POLYMER_ENTITY_INSTANCE",
    "BRANCHED_ENTITY_INSTANCE", "NONPOLYMER_ENTITY_INSTANCE",
    "CHEMICAL_COMPONENT"
  ),
  Typical_ID_Format = c(
    "4-character PDB ID, e.g. 4HHB",
    "Entry plus assembly ID, e.g. 4HHB-1",
    "Entry plus entity ID, e.g. 4HHB_1",
    "Entry plus branched entity ID",
    "Entry plus nonpolymer entity ID, e.g. 3PQR_5",
    "Instance or chain-level identifier, endpoint-specific",
    "Instance-level identifier, endpoint-specific",
    "Instance-level identifier, endpoint-specific",
    "Chemical component ID, e.g. ATP"
  ),
  Typical_Use = c(
    "Whole-structure metadata",
    "Biological assembly and symmetry",
    "Entity-level taxonomy or sequence annotations",
    "Glycan/branched entity records",
    "Ligand records within structures",
    "Chain-specific annotations",
    "Branched entity instance records",
    "Ligand instance records",
    "Ligand chemistry and descriptors"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(id_reference, align = c("l", "l", "l"))
```

The precise identifier syntax for instance-level records depends on the RCSB
schema and endpoint, but the key conceptual point is that the package supports
multiple biological levels and expects identifiers that match those levels.
The identifier helpers introduced in the current version make this easier to
manage explicitly: `infer_id_type()` classifies common patterns,
`parse_rcsb_id()` decomposes them, and the `build_*_id()` functions generate
normalized identifiers programmatically.

# Appendix D: Return Classes and Their Meaning

The package uses return classes as lightweight contracts. The most important
ones are summarized here.

```{r return-contract-table, results = "asis"}
contract_reference <- data.frame(
  Function = c(
    "query_search(return_type = 'entry')",
    "query_search(other return_type)",
    "perform_search()",
    "perform_search(return_with_scores = TRUE)",
    "perform_search(return_raw_json_dict = TRUE)",
    "fetch_data()",
    "data_fetcher_batch(return_as_dataframe = TRUE)",
    "data_fetcher(return_as_dataframe = TRUE)",
    "data_fetcher(return_as_dataframe = FALSE)",
    "as_rpdb_entry()",
    "as_rpdb_assembly()",
    "as_rpdb_polymer_entity()",
    "as_rpdb_chemical_component()",
    "as_rpdb_structure()"
  ),
  Return_Class = c(
    "rPDBapi_query_ids",
    "rPDBapi_query_response",
    "rPDBapi_search_ids",
    "rPDBapi_search_scores",
    "rPDBapi_search_raw_response",
    "rPDBapi_fetch_response",
    "rPDBapi_dataframe",
    "rPDBapi_dataframe",
    "rPDBapi_fetch_response",
    "rPDBapi_entry",
    "rPDBapi_assembly",
    "rPDBapi_polymer_entity",
    "rPDBapi_chemical_component",
    "rPDBapi_structure"
  ),
  Meaning = c(
    "Identifier vector from query_search()",
    "Parsed query_search payload",
    "Identifier vector from perform_search()",
    "Scored search results",
    "Raw JSON-like search payload",
    "Validated GraphQL fetch payload",
    "Flattened batch result with provenance metadata",
    "Flattened analysis-ready table",
    "Nested validated fetch payload",
    "Typed entry wrapper around retrieved data",
    "Typed assembly wrapper around retrieved data",
    "Typed polymer-entity wrapper around retrieved data",
    "Typed chemical-component wrapper around retrieved data",
    "Typed structure wrapper around retrieved data"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(contract_reference, align = c("l", "l", "l"))
```

These classes are useful when writing wrappers, tests, or pipelines that need
to branch on the kind of object returned by the package.

# Appendix E: Error and Failure-Mode Guidance

The package also uses typed errors in several important places. Users do not
need to memorize these classes for normal interactive work, but they are useful
for robust scripting and package development.

```{r error-guidance}
error_guidance <- data.frame(
  Scenario = c(
    "Malformed search response",
    "Unsupported return-type mapping",
    "Invalid input to search/fetch helper",
    "Unknown property or subproperty in strict mode",
    "Batch retrieval failure after retries",
    "HTTP failure",
    "Response parsing failure"
  ),
  Typical_Class_or_Source = c(
    "rPDBapi_error_malformed_response",
    "rPDBapi_error_unsupported_mapping",
    "rPDBapi_error_invalid_input",
    "validate_properties() / generate_json_query()",
    "data_fetcher_batch()",
    "handle_api_errors() / send_api_request()",
    "parse_response()"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(error_guidance, align = c("l", "l"))
```

In practice, these classes matter when you want to distinguish network failures
from schema mismatches or user-input errors. That distinction is particularly
important in automated structural bioinformatics pipelines that may run over
many identifiers.

# Reproducible Research with rPDBapi

Programmatic structure retrieval is most useful when the search logic and the
retrieved identifiers are stored alongside the analysis. A practical workflow is
to save:
  
  - the query object used for the search
- the vector of returned identifiers
- the selected metadata fields
- the property-validation mode and identifier-construction rules
- any batch provenance or cache configuration
- object-level metadata attached to typed wrappers
- the session information and package versions

```{r reproducibility}
analysis_manifest <- list(
  live_examples = live_examples,
  package_version = as.character(utils::packageVersion("rPDBapi")),
  query = kinase_query,
  requested_entry_fields = entry_properties,
  strict_property_validation = getOption("rPDBapi.strict_property_validation", FALSE),
  built_ids = list(
    entry = build_entry_id("4HHB"),
    assembly = build_assembly_id("4HHB", 1),
    entity = build_entity_id("4HHB", 1),
    instance = build_instance_id("4HHB", "A")
  ),
  batch_provenance_example = if (live_examples) {
    attr(kinase_batch, "provenance")
  } else {
    NULL
  }
)

str(analysis_manifest, max.level = 2)
```

This manifest is a simple example of how to preserve the logic of an analysis.
Because the search operators and requested fields are explicit R objects, they
can be saved with `saveRDS()` and reused later. That is a better long-term
strategy than relying on manual notes about which website filters were used.
When batch retrieval is part of the workflow, the provenance attribute from
`data_fetcher_batch()` provides an additional audit trail for how the data were
obtained.

# Summary

`rPDBapi` supports an end-to-end workflow for structural bioinformatics in R:
  search the archive, refine the result set with explicit operators, validate
properties against known schema fields, work across identifier levels,
retrieve entry-, entity-, and assembly-level metadata, scale retrieval with
batch and cache-aware helpers, convert nested responses into tidy data frames
or typed objects, download coordinate files, and integrate the results with
analysis and visualization packages. This workflow is useful not only for
exploratory access to the PDB, but also for reproducible, scriptable analyses
that can be revised and rerun as biological questions evolve.

# Session Information

```{r session-info}
sessionInfo()
```

Session information is included so that rendered results can be tied to a
specific R and package environment. This is a standard but important part of
reproducible computational biology documentation.
